---
title: "Mini Data-Analysis Deliverable 1"
output: html_document
---

# Welcome to your (maybe) first-ever data analysis project!

And hopefully the first of many. Let's get started:

1. Install the [`datateachr`](https://github.com/UBC-MDS/datateachr) package by typing the following into your **R terminal**:

```
install.packages("devtools")
devtools::install_github("UBC-MDS/datateachr")
```

2. Load the packages below.

```{r}
library(datateachr)
library(tidyverse)
```

3. Make a repository in the https://github.com/stat545ubc-2021 Organization. You will be working with this repository for the entire data analysis project. You can either make it public, or make it private and add the TA's and Vincenzo as collaborators. 
    - UPDATE: There is now a link that will help you make a repo. It's [on canvas](https://canvas.ubc.ca/courses/80816/pages/mini-data-analysis-create-a-repository) (and on a Slack announcement).

4. When you go to submit, submit a URL to your repository to canvas.  

# Instructions

More details regarding the instructions and points allocated to each task can be found below. Briefly,

+ Each milestone is worth 30 points. The number of points allocated to each task will be annotated within each deliverable. Tasks that are more challenging will often be allocated more points.

+ 10 points will be allocated to the reproducibility, cleanliness, and coherence of the analysis. While the three milestones will be submitted as independent deliverables, the analysis itself is a continuum - think of it as 3 chapters to a story. Each chapter, or in this case, portion of your analysis, should be easily followed through by someone unfamiliar with the content. [Here](https://swcarpentry.github.io/r-novice-inflammation/06-best-practices-R/) is a good resource for what constitutes "good code". Learning good coding practices early in your career will save you hassle later on!

# Learning Objectives

By the end of this milestone, you should:

+ Become familiar with your dataset of choosing
+ Think of 4 questions that you would like to answer with your data
+ Generate a reproducible and clear report using R Markdown

# Task 1: Choose your favorite dataset (10 points)

The `datateachr` package by Hayley Boyce and Jordan Bourak currently composed of 7 semi-tidy datasets for educational purposes. Here is a brief description of each dataset:

+ *apt_buildings*: Acquired courtesy of The City of Toronto’s Open Data Portal. It currently has 3455 rows and 37 columns.

+ *building_permits*: Acquired courtesy of The City of Vancouver’s Open Data Portal. It currently has 20680 rows and 14 columns.

+ *cancer_sample*: Acquired courtesy of UCI Machine Learning Repository. It currently has 569 rows and 32 columns.

+ *flow_sample*: Acquired courtesy of The Government of Canada’s Historical Hydrometric Database. It currently has 218 rows and 7 columns.

+ *parking_meters*: Acquired courtesy of The City of Vancouver’s Open Data Portal. It currently has 10032 rows and 22 columns.

+ *steam_games*: Acquired courtesy of Kaggle. It currently has 40833 rows and 21 columns.

+ *vancouver_trees*: Acquired courtesy of The City of Vancouver’s Open Data Portal. It currently has 146611 rows and 20 columns.

**Things to keep in mind**

+ We hope that this project will serve as practice for carrying our your own *independent* data analysis. Remember to comment your code, be explicit about what you are doing, and write notes in this markdown document when you feel that context is required. As you advance in the project, prompts and hints to do this will be diminished - it'll be up to you!

+ Before choosing a dataset, you should always keep in mind **your goal**, or in other ways, *what you wish to achieve with this data*. This mini data-analysis project focuses on *data wrangling*, *tidying*, and *visualization*. In short, it's a way for you to get your feet wet with exploring data on your own.

And that is exactly the first thing that you will do!

1.1 Out of the 7 datasets available in the `datateachr` package, choose **4** that appeal to you based on their description. Write your choices below:

```{r}
head(cancer_sample)
head(vancouver_trees)
head(steam_games)
head(apt_buildings)
```

**Note**: We encourage you to use the ones in the `datateachr` package, but if you have a dataset that you'd really like to use, you can include it here. But, please check with a member of the teaching team to see whether the dataset is of appropriate complexity. Also, include a **brief** description of the dataset here to help the teaching team understand your data.

1: CANCER_SAMPLE     
2: VANCOUVER_TREES   
3: STEAM_GAMES    
4: APT_BUILDINGS    

1.2 One way to narrowing down your selection is to *explore* the datasets. Use your knowledge of dplyr to find out at least *3* attributes about each of these datasets (an attribute is something such as number of rows, variables, class type...). The goal here is to have an idea of *what the data looks like*. 

*Hint:* This is one of those times when you should think about the cleanliness of your analysis. I added a single code chunk for you, but do you want to use more than one? Would you like to write more comments outside of the code chunk?

*Exploration notes* 
Glimpse() will show some information regarding columns and rows of the data; whereas head() will show the first few rows of the data to give an idea of the values I could potentially work with; lastly, I will also look at the summary() of the data for a basic overview. This will be repeated for all four data sets 

*exploring cancer_sample data set*

```{r}
glimpse(cancer_sample)
head(cancer_sample)
summary(cancer_sample)
```
*exploring vancouver_trees data set*
```{r}
glimpse(vancouver_trees)
head(vancouver_trees)
summary(vancouver_trees)
```

*exploring steam_games data set*
```{r}
glimpse(steam_games)
head(steam_games)
summary(steam_games)
```

*exploring apt_buildings data set* 
```{r}
glimpse(apt_buildings)
head(apt_buildings)
summary(apt_buildings)

```



1.3 Now that you've explored the 4 datasets that you were initially most interested in, let's narrow it down to 2. What lead you to choose these 2? Briefly explain your choices below, and feel free to include any code in your explanation. 

*CHOICE 1* cancer_sample: I am in biomedical research, so I was initially drawn to the data. Once I looked through the data briefly, I believe it could be interesting to see how the diagnosis itself could relate to several parameters and characteristics. This in turn could be analysed and could possibly used to train a program to support diagnoses. 

*CHOICE 2* vancouver_trees: I like this data set because it contains a good mixture of numerical and text variables, and because I believe it could be used to create a very detailed map/graph of distribution of trees at some point during the analysis. 

1.4 Time for the final decision! Going back to the beginning, it's important to have an *end goal* in mind. For example, if I had chosen the `titanic` dataset for my project, I might've wanted to explore the relationship between survival and other variables. Try to think of 1 research question that you would want to answer with each dataset. Note them down below, and make your final choice based on what seems more interesting to you!

*CHOICE 1* cancer_sample:how does texture and area influence the diagnosis of the sample, and how can we use this to build a predictive model in the future? 
*CHOICE 2* vancouver_trees:which city areas have the greatest concentration of planted trees? 


# Important note

Read Tasks 2 and 3 *fully* before starting to complete either of them. Probably also a good point to grab a coffee to get ready for the fun part!

This project is semi-guided, but meant to be *independent*. For this reason, you will complete tasks 2 and 3 below (under the **START HERE** mark) as if you were writing your own exploratory data analysis report, and this guidance never existed! Feel free to add a brief introduction section to your project, format the document with markdown syntax as you deem appropriate, and structure the analysis as you deem appropriate. Remember, marks will be awarded for completion of the 4 tasks, but 10 points of the whole project are allocated to a reproducible and clean analysis. If you feel lost, you can find a sample data analysis [here](https://www.kaggle.com/headsortails/tidy-titarnic) to have a better idea. However, bear in mind that it is **just an example** and you will not be required to have that level of complexity in your project.

# Task 2: Exploring your dataset (15 points)

If we rewind and go back to the learning objectives, you'll see that by the end of this deliverable, you should have formulated *4* research questions about your data that you may want to answer during your project. However, it may be handy to do some more exploration on your dataset of choice before creating these questions - by looking at the data, you may get more ideas. **Before you start this task, read all instructions carefully until you reach START HERE**.

2.1 Complete *4 out of the following 8 exercises* to dive deeper into your data. All datasets are different and therefore, not all of these tasks may make sense for your data - which is why you should only answer *4*. Use *dplyr* and *ggplot*.

2.2 For each of the 4 exercises that you complete, provide a *brief explanation* of why you chose that exercise in relation to your data (in other words, why does it make sense to do that?), and sufficient comments for a reader to understand your reasoning and code.

*I will be using the cancer data set (CHOICE 1). Some information about the data:*

- The data set comes from the Breast Cancer Wisconsin (Diagnostic) Data Set
- "Features are computed from a digitized image of a fine needle aspirate (FNA) of a breast mass. They describe characteristics of the cell nuclei present in the image" 
- containing the following attributes: 
1) ID number
2) Diagnosis (M = malignant, B = benign)
3-32)

Ten real-valued features are computed for each cell nucleus:

a) radius (mean of distances from center to points on the perimeter)
b) texture (standard deviation of gray-scale values)
c) perimeter
d) area
e) smoothness (local variation in radius lengths)
f) compactness (perimeter^2 / area - 1.0)
g) concavity (severity of concave portions of the contour)
h) concave points (number of concave portions of the contour)
i) symmetry
j) fractal dimension ("coastline approximation" -1)

*I will be completing the following 4 exercises*

1. Plot the distribution of a numeric variable.

I want to see how the area is distributed, and group it by diagnosis, to see what tumor sizes I can expect to see when I perform more analyses. 

```{r}
sizing_plot <- ggplot(cancer_sample, aes(area_mean)) +
  geom_bar(aes(color = diagnosis))

print(sizing_plot)

```
It looks like most higher values are going to be malignant, whereas most benign tumors tend to be smaller, although there is an area of overlap. We will have to look at other values to determine how we can differentiate between benign and malignant tumors. 


4. Explore the relationship between 2 variables in a plot.
Seeing some trend with the area, I also want to see how the area compares with the symmetry, and if there are any trends when looking at diagnoses. 

```{r}
area_symmetry_plot <- ggplot(cancer_sample, aes(area_mean, symmetry_mean))+ 
  geom_point(aes(colour = diagnosis))

print(area_symmetry_plot)
```
There seems to be a tighter knit cluster for benign tumors than malignant tumors. Will need to use other factors also. 


5. Filter observations in your data according to your own criteria. 

I want to try looking specifically at the radius, perimeter, and area. Seeing as mean area tends to be at most 1000 for benign values, I want to filter and see how other values look like when I filter for high area tumors, and see how many there are total 


```{r}
cancer_area <- cancer_sample %>% 
  filter(area_mean > 1000)

head(cancer_area)
```


2. Create a new variable based on other variables in your data (only if it makes sense)

I want to calculate the difference between the area_worst and area_mean? I am considering using this for the next analysis. 
```{r}

```


6. Use a boxplot to look at the frequency of different observations within a single variable. You can do this for more than one variable if you wish!


# Task 3: Write your research questions (5 points)

So far, you have chosen a data set and gotten familiar with it through exploring the data. Now it's time to figure out 4 research questions that you would like to answer with your data! Write the 4 questions and any additional comments at the end of this deliverable. These questions are not necessarily set in stone - TAs will review them and give you feedback; therefore, you may choose to pursue them as they are for the rest of the project, or make modifications!

1. How is perimeter, radius and area all together related to diagnose a tumor as malignant or benign?
2. Which features are the driving factor for the diagnosis?
3. What effects does symmetry have on the radius and area of the sample? 
4. Looking at the worst factors - how are these different in the diagnosis of the tumor compared to the mean values? 

### Attribution

Thanks to Icíar Fernández Boyano for mostly putting this together, and Vincenzo Coia for launching.
